\documentclass[a4paper, 10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[english]{babel}

\usepackage[
  a4paper,
  left=1.5cm,
  right=1.5cm,
  top=1cm,
  bottom=2cm
  ]{geometry}
  
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm} 
\usepackage{thmtools}
\usepackage{enumitem}

\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{Midterm Exam: Grondslagen van de Wiskunde}
\date{December 16, 2025, 09:00-12:00}

\begin{document}
\maketitle

\noindent \textsc{This exam consists of 5 exercises; see also the
  back side.}

\noindent Every exercise is worth 10 points; your grade is the total
divided by 5. If an exercise contains several questions, the number of
points attributed to each question is indicated.

\noindent As the exam is \textbf{closed book}, some basics definition
about posets, which you may use freely, are recalled below.

\noindent Advice: start first with the exercises that you can do, and
then think about the rest. Succes!

\medskip

\hrule

\medskip

\noindent\textit{A few definitions about posets:} Recall that a poset is a pair
$(X,\leq)$ of a set $X$ and a binary relation $\leq$ such that:
\begin{itemize}
\item (reflexivity) $x\leq x$ for any $x\in X$
\item (anti-symmetry) $x\leq y$ and $y \leq x$ imply $x = y$ for any
  $x,y \in X$, 
\item (transitivity) $x\leq y$ and $y \leq z$ imply $x\leq z$, for any
  $x,y,z \in X$.
\end{itemize}
 We say furthermore that the order is \emph{linear} (or \emph{total}) if
for any $x,y \in X$, we have $x \leq y$ or $y \leq x$.

\noindent We will often only refer to a poset as $X$ instead of
$(X,\leq)$.

\noindent A poset is \emph{well-ordered} if every non-empty subset has
a least element.

\noindent A function $f \colon L \to M$ between well-ordered sets is
an \emph{embedding} if it is strictly increasing (see Exercise 1) and
for any $x \in L$ and any $y\in M$ such that $y \leq f(x)$, there
exists $x' \leq x$ with $f(x')=y$.

\medskip

\hrule

\medskip 

\medskip

\noindent\textbf{Exercise 1.} 

  \begin{enumerate}
  \item (2 points) Let $P$ and $Q$ be posets and $f \colon P \to Q$ a
    function. Prove that if $P$ is linearly ordered, then the following
    assertions are equivalent:
    \begin{enumerate}[label=(\alph*)]
    \item $f$ is order-preserving (if $x \leq y$ then $f(x)\leq f(y)$)
      and injective,
    \item $f$ is strictly increasing: if $x<y$ then $f(x)<f(y)$,
    \item $f$ satisfies the condition: $x<y$ if and only if $f(x)<f(y)$.
    \end{enumerate}
  \item (4 points) Let $L$ and $M$ two well-ordered sets, $f \colon L \to M$ an
    embedding and $g \colon L \to M$ a strictly increasing
    function. Prove that for every $x \in L$, we have
    \[
      f(x)\leq g(x).
    \]
  \item (4 points) Let $L$ and $M$ be two well-ordered sets, prove that there
    exist an embedding $L \to M$ if and only if there exist a strictly
    increasing function $L \to M$.
  \end{enumerate}

  \medskip

  
\noindent\textbf{Exercise 2.} A \emph{tree} is a poset $(T,\leq)$ such
that:
\begin{enumerate}[label=(\roman*)]
\item it has a least element, called the \emph{root},
\item for every
  $x \in T$, the set $\{y \in T\,\vert\, y \leq x\}$ is well-ordered.
\end{enumerate}
Given an element $x \in T$, an \emph{immediate successor} of $x$ is an
element $y$ such that $x < y$ and there is no $z$ such that $x < z <
y$.
We say that a tree is \emph{infinite} if the underlying set is
infinite, and we say that it is \emph{locally finite} if for every
element $x \in T$, the set of immediate successors of $x$ is finite.
\begin{enumerate}
\item (3 points) Let $(T,\leq)$ be an infinite tree which is locally
  finite. Prove that if $x$ is an element such that the set  $\{y \in
  T\,\vert\, y \geq x\}$ is infinite, then there is an immediate
  successor of $x$ having the same property.
\item (7 points) Prove the famous K\H{o}nig's lemma: For every infinite and locally
  finite tree, there exists an infinite strictly increasing
  sequence
  \[
    x_0 < x_1 < x_2 < \dots.
  \]
  You have to make explicit where you use the axiom of choice in your answer.
\end{enumerate}

\medskip

\noindent\textbf{Exercise 3.} Let $X$ be an \emph{infinite} set and let $A
\subseteq X^X$ an infinite set of functions $X \to X$. Let
$\overline{A}$ be the smallest subset of $X^X$ such that: $A \subseteq
\overline{A}$ and $\overline{A}$ is closed under composition, i.e. if
$f,g \in \overline{A}$.

Show that $\vert \overline{A} \vert = \vert A \vert$.

\medskip

\noindent\textbf{Exercise 4.} Consider the language $L=\{f\}$, with $f$
is a unary function symbol, and the $L$-structure $R$, whose
underlying set is $\R$, and $f^R(x)=x^2$.
\begin{enumerate}
\item (5 points) Using $L$-formulas, define in $R$ the
  numbers $-1, 0$ and $1$. That is, give $L$-formulas with one free
  variable such that the set of elements of $R$ for which the formula
  holds is $\{-1\}$ (resp.\ $\{0\}$, resp.\ $\{1\}$). 
\item (5 points) Find an $L$-formula $\phi(x)$ such that the subset of $x \in
  R$ that satisfies $\phi(x)$ is the subset $\{x \in \R \,\vert\, x> 0\}$.
\end{enumerate}

% \noindent\textbf{Exercise 4.} For each of the following set,
% determine if they are finite, infinite countable or
% uncountable. Justify your answers.
% \[
%   \begin{aligned}
%    \text{(4 points)}\quad A&=\{f \colon \N \to \N \,\vert\, f(n)=f(n+1)+f(n+2)\}\\
%     \text{(3 points)}\quad B&=\{f \colon \N \to \N \,\vert\, f(n)^2-3f(n)+2=0\}\\
%     \text{(3 points)}\quad C&=\{f \colon \N \to \N \,\vert\, f(n)\geq f(n+1)\}
                                %   \end{aligned}
                                %                                 \]


\medskip

\noindent\textbf{Exercise 5.} Let $(X,\leq)$ be a poset. The goal of
the exercise is to prove that there always exists a linear order $\preceq$ on
$X$ that extends $\leq$ (i.e.\ $x \leq y$ implies $x\preceq y$).
\begin{enumerate}
\item (4 points) Prove the statement in the case that $X$ is finite.
\item (6 points) Prove the general case using the compactness theorem.
\end{enumerate}


\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
